`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models

### 2025-10-07 paper/paper.qmd
edits shauna

In global health estimation, Bayesian hierarchical models have become a cornerstone methodology, enabling the integration of diverse data sources to produce reliable estimates across multiple populations and regions. For example, they are used to estimate family planning indicators for the FP2030 initiative [@alkema_2024evol], incidence of low birth weight [@okwaraji_national_2024], abortion and unintended pregnancies rates [@bearak_unintended_2020], and maternal mortality by the United Nations Interagency Group [@peterson_bayesian_2024]. In these efforts, national or subnational estimates are often required for population-periods with limited data availability. Bayesian hierarchical models—whether used directly or for specific components within larger modeling frameworks—are valuable in these contexts because they facilitate the sharing of information across populations. As a result, global modeling exercises typically fit Bayesian hierarchical models to comprehensive global databases, producing national-level estimates for all countries worldwide.

@@

This paper introduces the \texttt{localhierarchy} R package designed to address these challenges by supporting the development and fitting of Bayesian hierarchical models in settings requiring both global and local estimation. The package provides R functions and Stan functions and model blocks <!-- citep{stan2023}  --> to facilitate specification of hierarchical models, to fit full global models, and to subsequently fit local models using fixed parameter values inferred from the global analysis. Through this framework, users can conveniently perform local inference informed by global context, streamlining workflows in global health estimation tasks.

@@

# The Family Planning Estimation Tool (FPET)

The Family Planning Estimation Tool (FPET) is widely used to estimate key family planning indicators in countries and regions. Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to the ratio of modern use to demand. Data on such indicators is available from household surveys and routine data systems. FPET estimates are obtained from fitting a Bayesian model to available data. An illustration of FPET estimates and available data is shown in @fig-localfits for three countries in Western Africa.

@@

Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in Figure \ref{fig-hierstruc}. Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the asymptote of the transition of demand from low to high values, and the level of demand in a reference year. @fig-level shows estimates of the level in the reference year at the country level for countries in Western Africa, and for the groupings used in the hierarchical model. @fig-asympt shows estimates for the asymptote. These figures show variation in levels among countries in Western Africa, informed by country-specific data. Estimates of asymptotes are similar across countries in Western Africa because countries have not yet reached the end phase of their transition; the asymptotes in Western African countries are informed by the estimate of the asymptote in sub-Saharan Africa through the hierarchical model.

@@

![Estimates of the level of demand for countries in Western Africa and for higher levels used in the hierarchical model (smaller world regions and major world regions). Point estimates and 95% credible intervals are shown.](fig/hier_level.jpg){#fig-level width="685"}

![Estimates of the asymptote for demand for countries in Western Africa and for higher levels used in the hierarchical model (major world regions). Point estimates and 95% credible intervals are shown.](fig/hier_asymptote.jpg){#fig-asympt}

The workflow used in FPET to produce country-level estimates is as follows. Firstly, the full Bayesian model is fitted to a global data base of survey data. Data base compilation is carried out by the United Nations Population Division (see [UNPD world contraceptive use website](https://www.un.org/development/desa/pd/world-contraceptive-use)) and the Track20 project (see [www.track20.org](https://www.track20.org/)). The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes. Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit. The technical details for this approach are given in the next section. We return to the application of estimating family planning indicators in Section 4.2.

@@

Suppose we want to estimate a population parameter $\mu_r^{(l)}$ of interest, referring to a population health indicator for region $r$ at level $l$ of the hierarchy. For example, the indicator could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes. For a given level $l$, index $r$ refers to the population in a particular region indexed by $r$, for example, a subnational region, a country, or a group of countries. We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 6: \begin{align*}
  \mu_{r}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma_l &\sim N(\mu^{(l-1)}_{r^\prime}, \sigma_l^2), \text{ for } l > 0,
\end{align*} where $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$. The parameter $\sigma_l$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy. At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.

@@

% \draw[arrow] (mu31.south) -- (mu41.north);
% \draw[arrow] (mu31.south) -- (mu42.north);
% \draw[arrow] (mu31.south) -- (dots4.north);
\end{tikzpicture}
\end{adjustbox}
\caption{General specification of a hierarchical model.}
\end{figure}
```

As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure 2). In this setting, using alternative notation for the same setup to indicate the level by name, the hierarchical model can be written as follows:

@@

We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set. A typical workflow for local estimation of national level outcomes is illustrated in Figure 7. For national estimation using the model given by Eq.\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model. In a local model, data from one country are used and parameters that are not country-specific are fixed using parameter estimates from the global model fit.

@@

% --- Legend ---
% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[below=1cm of localmodel](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
\end{tikzpicture}
\end{adjustbox}
\caption{Workflow for global and local national modeling. }
\end{figure}
```

@@

replacing the hierarchical mean and variances by point estimates. Details are given in the implementation section.

The `localhierachy` package provides a general implementation of local models by facilitating the fixing of subsets of parameters. Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma^{(k)}$ for $k=1, ..., l$ and fixing $\mu_{r\prime}{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit. More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma^{(k)}$ at all levels $k$ up to some chosen level $K_{\sigma}$. Similarly, the $\mu^{(k)}$ parameters can be fixed at all levels $k$ up to a level $K_{\mu} \leq K_{\sigma}$.

## Implementation

@@

In a typical work flow, a user first fits a global model, followed by local ones. Worked examples are given in the use case section. The global and local models in these examples are fitted using the `localhierarchy` wrapper function `fit_model_localhierarchy`, using its arguments to specify a hierarchical model, data, and, for local fits, pass on outputs from the global model. The wrapper function calls R functions to process data and prepare Stan input data, read in a Stan model, fit the model, and to postprocess the results.

The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs. A worked example is given in the first use case. Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c^{(\country)} =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:

@@

The first level is the intercept, which is the global mean, followed by the region, subregion, and country levels. The user can then specify a local model for one country by fixing parameters at different levels, using the `hierarchical_sigmas_fixed` and `hierarchical_terms_fixed` arguments in the `fit_model_localhierarchy` function. For example, if the $\sigma$s are to be fixed up to the country level and if the $\mu$ terms are to be fixed up to the subregion, the user can encode this as follows:

@@

Implementation is based on a model matrix. For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c^{(\country)}$s: $$(\mu_1^{(\country)}, ..., \mu_c^{(\country)}) = M\cdot (\eta_1, ..., \eta_{1+R+S+C})',$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\eta_1 = s_{\glob}\cdot \gamma_{\glob}, \eta_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)}, ..., \eta_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}^{(\country)}$. When estimating a subset of parameters in a local model, the values of $\sigma$s and $\gamma$s are fixed to point estimates (posterior means) from a global fit. For example, for a national model, we get \begin{align*}
  \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
          \hat{\sigma}_{ \region} \cdot \hat{\gamma}_{w[s[c]]]}^{(\region)} + 
          \hat{\sigma}_{ \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
          \hat{\sigma}_{ \country} \cdot \gamma_{c}^{(\country)}.
\end{align*}

@@

# Use case: National estimation

@@

The simulation setup is summarized in Figure 8, panels A and B. Simulation of subnational data is discussed in a later section.

@@

% --- Arrows ---
\draw[arrow] (mu0.south) -- (mu11.north);
\draw[arrow] (mu0.south) -- (mu12.north);
\draw[arrow] (mu0.south) -- (dots1.north);
\end{tikzpicture}
\end{adjustbox}}
\caption{Diagram of simulation settings. Top: Hierarchical model for generating country and subnational parameters. Center: Data generating mechanism for national data, illustrated for the first country. Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}
\end{figure}
\clearpage
```

Panel A in Figure 4 indicates the hierarchical model considered. For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c^{(\country)}|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$. Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}^{(\country)}, \sigma_y^2 \sim N(\mu_{c[i]}^{(\country)}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$. We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling. In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.

@@

Elements include the model matrix and other elements are used to make a link between columns in the model matrix, or equivalently, indices in the vector $\eta$, and the hierarchical groupings. The model matrix $M$, as introduced above, is as follows:

@@

All this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting. This function outputs a named list with Stan input data relevant to the hierarchical setup:

@@

We can use the function `fit_model_localhierarchy` to fit the model as specified so far. The argument `runstep` is used to define the type of model, setting it to `"global_national"` results in fitting a model to all data without fixing any parameters. Other arguments include the hierarchical level specification, the area, and survey data:

@@

A local fit can be esetimated by passing on the global fit object to the wrapper function and including data for countries of interest. Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:

@@

There are several functions to produce and check outputs. The function `posterior_summary_hierparam` is used to summarize posterior draws for hierarchical parameters, to produce summaries of $\mu$ at any level of interest. It takes a fitted model object and a parameter name as input and returns a list with each entry referring one level, and the results (e.g., estimates for $\mu$ estimates) for that level. Here we obtain these results for the global and local model:

@@

The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:

@@

## Application: National estimation of family planning indicators using FPET

@@

For local estimation of family planning indicators, the Bayesian model described in section 2 is used. Specifically, hierarchical models are used for estimation of transition model parameters. The workflow used for estimating family planning indicators at the national level is based on model fitting in several steps, including the estimation of long-term trends in the first step, the estimation of all model parameters related to modern use in a second step, the addition of information on the use of traditional methods of contraception in a third step, followed by local model fitting to produce estimates for all indicators of interest. The `localhierarchy` functionality is incorporated in the family planning modeling code base such that all steps are run with the same code base, and parameters can be passed from one model fit to another. Full details on this workflow and model specification are described elsewhere [@alkema_2024evol]. As an illustration, @fig-allfits shows global and local fits for modern use, demand, and demand satisfied for selected countries in Western Africa. The global fit in red corresponds closely to the local fits, shown in green.

@@

# Use case 2: Subnational estimation

@@

For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level. If data are limited, models can be considered to share information across regions. We consider here the use of hierarchical models for an indicator of interest. The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach. Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions, such as the one from Figure 3. When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.

@@

The approach followed in NISE is illustrated in Figure 4, for the hierarchical structure from Figure 2. In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters. This national-level model fit is used in a global subnational fit to a global data base of subnational data. Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{ l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level. Finally, in a 3rd step, local models can be fitted, using fixed values from the global subnational fit. In the remainder of this section, we illustrate this approach using simulated data, and present an application of this approach for family planning estimation.

@@

% --- Legend ---
% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[below=1.2cm of localmodel.west](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
\end{tikzpicture}
\end{adjustbox}
\caption{Workflow for nationally-informed global and local subnational modeling. }
\end{figure}
\clearpage
```

@@

The simulation setup for subnational parameters and data is summarized in Figure 8, Panels A-B-C. As explained above for national estimation, Panel A indicates the hierarchical model considered. For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]}^{(\country)} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{ \subnat} \sim N(0, \sigma_{ \subnat})$.

As explained earlier as well, we chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling. Subnational data follows a normal distribution with constant variance: $$y^{(subnat)}_i|\mu^{(subnat)}_{r[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $r[i]$ refers to the subnational region of subnational observation $i$. In this simulation setup, for simplicity, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels. In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source. We discuss an example of those settings in the case study.

@@

Figure @fig-subnatfits show illustrations of regional fits for Benin. The subnational model is fitted to all regional data from the country. @fig-subnatoverview illustrates fits in selected regions and differences across regions in 2024.

@@

We have used the `localHierarchy` functionality in different projects. We discussed the estimation of family planning indicators, where country M&E officers use the local model to produce estimates at the national and subnational level. In this area, M&E officers have successfully used local models to produce national estimates since 2014. Availability of a local modeling tool has enabled users to do additional analyses and check sensitivity of estimates to specific data sources or data quality assumptions. In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems (e.g., the number of contraceptive commodities supplied, see [@mooney_emu]). These use cases illustrate the value of local models to enable local analysis.

@@

>>>>

In global health estimation, Bayesian hierarchical models have become a cornerstone methodology, enabling the integration of diverse data sources to produce reliable estimates across multiple populations and regions. For example, they are used to estimate and forecast family planning indicators for the FP2030 initiative [@alkema_2024evol], incidence of low birth weight [@okwaraji_national_2024], abortion and unintended pregnancies rates [@bearak_unintended_2020], and maternal mortality by the United Nations Interagency Group [@peterson_bayesian_2024]. In these efforts, national or subnational estimates are often required for population-periods with limited data availability. Bayesian hierarchical models—whether used directly or for specific components within larger modeling frameworks—are valuable in these contexts because they facilitate the sharing of information across populations. As a result, global modeling exercises typically fit Bayesian hierarchical models to comprehensive global databases, producing national-level estimates for all countries worldwide.

@@

This paper introduces the \texttt{localhierarchy} R package designed to address these challenges by supporting the development and fitting of Bayesian hierarchical models in settings requiring both global and local estimation. The package provides R functions and Stan functions and model blocks [@stan2025] to facilitate specification of hierarchical models, to fit full global models, and to subsequently fit local models using fixed parameter values inferred from the global analysis. Through this framework, users can conveniently perform local inference informed by global context, streamlining workflows in global health estimation tasks.

@@

# The Family Planning Estimation Tool (FPET) {#sec-fpet}

The Family Planning Estimation Tool (FPET) is widely used to estimate key family planning indicators in countries and regions. Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to the ratio of modern contraceptive use to demand. Data on such indicators is available from household surveys and routine data systems. FPET estimates are obtained from fitting a Bayesian model to available data. An illustration of FPET estimates and available data is shown in @fig-localfits for three countries in Western Africa.

@@

Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in Figure \ref{fig-hierstruc}. Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the level of demand in a reference year and the asymptote of the transition of demand from low to high values. @fig-level shows estimates of the level in the reference year at the country level for countries in Western Africa, and for the groupings used in the hierarchical model. @fig-asympt shows estimates for the asymptote. These figures show variation in levels among countries in Western Africa, informed by country-specific data. Estimates of asymptotes are similar across countries in Western Africa because countries have not yet reached the end phase of their transition; the asymptotes in Western African countries are informed by the estimate of the asymptote in sub-Saharan Africa through the hierarchical model.

@@

![Estimates of the level of demand for major world regions (top), smaller world regions (middle) and countries in Western Africa (bottom). Point estimates and 95% credible intervals are shown.](fig/hier_level.jpg){#fig-level width="685"}

![Estimates of the asymptote for demand for major world regions (top) and countries in Western Africa (bottom). Point estimates and 95% credible intervals are shown.](fig/hier_asymptote.jpg){#fig-asympt}

The workflow used in FPET to produce country-level estimates is as follows. Firstly, the full Bayesian model is fitted to a global data base of survey data. Data base compilation is carried out by the United Nations Population Division (see [UNPD world contraceptive use website](https://www.un.org/development/desa/pd/world-contraceptive-use)) and the Track20 project (see [www.track20.org](https://www.track20.org/)). The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes. Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit. The technical details for this approach are given in the next section. We return to the application of estimating family planning indicators in @sec-fpetnat.

@@

Suppose we want to estimate a population parameter $\mu_r^{(l)}$ of interest, referring to a population health indicator for region $r$ at level $l$ of the hierarchy. For example, the indicator could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes. For a given level $l$, index $r$ refers to the population in a particular region indexed by $r$, for example, a subnational region, a country, or a group of countries. We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure \ref{fig-hiermodel}:

\begin{align*}
  \mu_{r}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma_l &\sim N(\mu^{(l-1)}_{r^\prime}, \sigma_l^2), \text{ for } l > 0,
\end{align*} where $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$. The parameter $\sigma_l$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy. At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.

@@

% \draw[arrow] (mu31.south) -- (mu41.north);
% \draw[arrow] (mu31.south) -- (mu42.north);
% \draw[arrow] (mu31.south) -- (dots4.north);
\end{tikzpicture}
\end{adjustbox}
\caption{General specification of a hierarchical model.}\label{fig-hiermodel}
\end{figure}
```

As a specific example, we consider the estimation of country-specific parameters, where countries are organized into subregions of countries (illustrated in Figure \ref{fig-hiermodel}). In this setting, using alternative notation for the same setup to indicate the level by name, the hierarchical model can be written as follows:

@@

We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set. A typical workflow for local estimation of national level outcomes is illustrated in Figure \ref{fig-workflow}. For national estimation using the model given by Eq.\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model. In a local model, data from one country are used and parameters that are not country-specific are fixed using parameter estimates from the global model fit.

@@

% --- Legend ---
% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[below=1cm of localmodel](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
\end{tikzpicture}
\end{adjustbox}
\caption{Workflow for global and local national modeling. }\label{fig-workflow}
\end{figure}
```

@@

replacing the hierarchical mean and variances by point estimates $\hat{\mu}_{s[c]}^{(\subregion)}$ and $\hat{\sigma}_{ \country}^2$, respectively. Details are given in the implementation section.

The `localhierachy` package provides a general implementation of local models by facilitating the fixing of subsets of parameters. Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma_k$ for $k=1, ..., l$ and fixing $\mu_{\cdot}^{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit. More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma_k$ at all levels $k$ up to some chosen level $K_{\sigma}$. Similarly, the $\mu^{(k)}$ parameters can be fixed at all levels $k$ up to a level $K_{\mu} \leq K_{\sigma}$.

## Implementation {#sec-implementation}

@@

In a typical work flow, a user first fits a global model, followed by local ones. Worked examples are given in the use cases in @sec-nationalusecase and @sec-subnatusecase. The global and local models in these examples are fitted using the `localhierarchy` wrapper function `fit_model_localhierarchy`, using its arguments to specify a hierarchical model, data, and, for local fits, pass on outputs from the global model. The wrapper function calls R functions to process data and prepare Stan input data, read in a Stan model, fit the model, and to postprocess the results.

The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs. A worked example is given in the first use case (@sec-nationalusecase). Briefly, if a user wants to specify a hierarchical model with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c^{(\country)} =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows:

@@

The first level is the intercept, which is the global mean, followed by the region, subregion, and country levels. The user can then specify a local model for one country by fixing parameters at different levels, using the `hierarchical_sigmas_fixed` and `hierarchical_terms_fixed` arguments in the `fit_model_localhierarchy` function. For example, if the $\sigma$ parameters are to be fixed up to the country level and if the $\mu$ terms are to be fixed up to the subregion, the user can encode this as follows:

@@

Implementation is based on a model matrix. For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c^{(\country)}$s: $$(\mu_1^{(\country)}, ..., \mu_c^{(\country)}) = M\cdot (\eta_1, ..., \eta_{1+R+S+C})',$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$ terms, i.e., $\eta_1 = s_{\glob}\cdot \gamma_{\glob}, \eta_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)}, ..., \eta_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}^{(\country)}$. When estimating a subset of parameters in a local model, the values of $\sigma$s and $\gamma$s are fixed to point estimates (posterior means) from a global fit. For example, for a national model, we get \begin{align*}
  \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
          \hat{\sigma}_{ \region} \cdot \hat{\gamma}_{w[s[c]]]}^{(\region)} + 
          \hat{\sigma}_{ \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
          \hat{\sigma}_{ \country} \cdot \gamma_{c}^{(\country)}.
\end{align*}

@@

# Use case: National estimation {#sec-nationalusecase}

@@

The simulation setup is summarized in Figure \ref{fig-simusetup}, top and center panels. Simulation of subnational data is discussed in a later section.

@@

% --- Arrows ---
\draw[arrow] (mu0.south) -- (mu11.north);
\draw[arrow] (mu0.south) -- (mu12.north);
\draw[arrow] (mu0.south) -- (dots1.north);
\end{tikzpicture}
\end{adjustbox}}
\caption{Diagram of simulation settings. Top: Hierarchical model for generating country and subnational parameters. Center: Data generating mechanism for national data, illustrated for the first country. Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}\label{fig-simusetup}
\end{figure}
\clearpage
```

The top panel in Figure \ref{fig-simusetup} indicates the hierarchical model considered. For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c^{(\country)}|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$. The center panel introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}^{(\country)}, \sigma_y^2 \sim N(\mu_{c[i]}^{(\country)}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$. We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling. In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.

@@

Elements include the model matrix and other elements are used to make a link between columns in the model matrix, or equivalently, indices in the vector $\eta$, and the hierarchical groupings. The model matrix $M$, as introduced in @sec-implementation, is as follows:

@@

All of this information is passed to the function `hierarchical_param_stan_data` to produce inputs needed in the Stan model fitting. This function outputs a named list with Stan input data relevant to the hierarchical setup:

@@

We can use the function `fit_model_localhierarchy` to fit the model as specified so far. Details are included in the appendix (@sec-helpfilefitmodel). The argument `runstep` is used to define the type of model, setting it to `"global_national"` results in fitting a model to all data without fixing any parameters. Other arguments include the hierarchical level specification, the area, and survey data. When calling the function, print statements are used to inform the user of key points related to the model fit:

@@

A local fit can be estimated by passing on the global fit object to the wrapper function and including data for countries of interest. Here we do local runs for all countries, by including the entire data set, and specify what to fix using the `runstep` argument:

@@

The package includes several functions to produce and check outputs. The function `posterior_summary_hierparam` is used to summarize posterior draws for hierarchical parameters, to produce summaries of $\mu$ at any level of interest. It takes a fitted model object and a parameter name as input and returns a list with each entry referring one level, and the results (e.g., estimates for $\mu$ estimates) for that level. Here we obtain these results for the global and local model:

@@

The column `name` includes the country names, given by simple indicares in this simulation. The column `quant` above refers to posterior means and upper and lower bounds of 95% credible intervals. We can create plots to show outputs, for example for the first 4 countries:

@@

## Application: National estimation of family planning indicators using FPET {#sec-fpetnat}

@@

For local estimation of family planning indicators, the Bayesian model described in @sec-fpet is used. Specifically, hierarchical models are used for estimation of transition model parameters. The workflow used for estimating family planning indicators at the national level is based on model fitting in several steps, including the estimation of long-term trends in the first step, the estimation of all model parameters related to modern contraceptive use in a second step, the addition of information on the use of traditional methods of contraception in a third step, followed by local model fitting to produce estimates for all indicators of interest. The `localhierarchy` functionality is incorporated in the family planning modeling code base such that all steps are run with the same code base, and parameters can be passed from one model fit to another. Full details on this workflow and model specification are described elsewhere [@alkema_2024evol]. As an illustration, @fig-allfits shows global and local fits for modern contraceptive use, demand, and demand satisfied for selected countries in Western Africa. The global fit in red corresponds closely to the local fits, shown in green.

@@

# Use case 2: Subnational estimation {#sec-subnatusecase}

@@

For many global health applications, subnational estimation is of interest to inform policy and programmatic decision making at a more local level. If data are limited, models can be considered to share information across regions. We consider here the use of hierarchical models for an indicator of interest. The `localhierarchy` package can be used to fit subnational hierarchical models, following the same principles as for national estimation or through an updated approach. Just as for national-level hierarchical modeling, the starting point for subnational hierarchical modeling is to specify a hierarchical modeling structure for the subnational regions. When following the same principles as for national-level hierarchical modeling, a global model can be fitted to a global data base of subnational data, followed by local models for subnational regions.

@@

The approach followed in NISE is illustrated in Figure \ref{fig-nise}. In the NISE approach, the first step is to fit a model to a global data base with national-level data, to obtain estimates of model parameters up to and including national level parameters. This national-level model fit isthen used in a global subnational fit to a global data base of subnational data. Specifically, in the global subnational fitting step, national level parameters are fixed at point estimates from the global national fit, i.e., $\hat{\sigma}_{ l}$ and $\mu_{\cdot}^{(l)}$ for level $l$ up to the country level. Finally, in a third step, local models can be fitted, using fixed values from the global subnational fit. In the remainder of this section, we illustrate this approach using simulated data, and present an application of this approach for family planning estimation.

@@

% --- Legend ---
% \node[arrowparam, below=1.2cm of localmodel.west, anchor=west] (leg1) {}; 
% \node[right=0.3cm of leg1] {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[below=1.2cm of localmodel.west](parameter) {\textcolor{param}{$\rightarrow$ Parameter estimates}};
\node[right=1cm of parameter] {\textcolor{data}{$\rightarrow$ Data}};
\end{tikzpicture}
\end{adjustbox}
\caption{Workflow for nationally-informed global and local subnational modeling. }\label{fig-nise}
\end{figure}
\clearpage
```

@@

The simulation setup for subnational parameters and data is summarized in Figure \ref{fig-simusetup} in three panels. As explained above for national estimation, the top panel indicates the hierarchical model considered. For subnational estimation, an additional level is added, with $\mu_r^{(\subnat)} = \mu_{c[r]}^{(\country)} + \eta_{r}^{(\subnat)}$, $\eta_{r}^{(\subnat)}|\sigma_{ \subnat} \sim N(0, \sigma_{ \subnat})$.

As previously explained, we chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling. Subnational data follows a normal distribution with constant variance: $$y^{(subnat)}_i|\mu^{(subnat)}_{r[i]}, \sigma_y^2 \sim N(\mu^{(subnat)}_{r[i]}, \sigma_y^2),$$ where $r[i]$ refers to the subnational region of subnational observation $i$. In this simulation setup, for simplicity, we generate subnational data independently of national-level data, and we use the same constant data variance at the national and subnational levels. In real examples, data variances are likely to differ (for example, because subnational sampling errors are larger at the subnational level) and national and subnational data may be derived from the same source. We discuss an example of those settings in the case study.

@@

@fig-subnatfits show illustrations of regional fits for Benin. The subnational model is fitted to all regional data from the country. @fig-subnatoverview illustrates fits in selected regions and differences across regions in 2024.

@@

We have used the `localHierarchy` functionality in different projects. We discussed the estimation of family planning indicators, where country M&E officers use the local model to produce estimates at the national and subnational level. In this area, M&E officers have successfully used local models to produce national estimates since 2014. Availability of a local modeling tool has enabled users to do additional analyses and check sensitivity of estimates to specific data sources or data quality assumptions. In addition to fitting the model to survey data, users of the local family planning model can also incorporate additional data from routine information systems (e.g., the number of contraceptive commodities supplied at family planning facilities, see [@mooney_emu]). These use cases illustrate the value of local models to enable local analysis.

@@

## Appendix II: Documentation of the `fit_model_localhierarchy` function {#sec-helpfilefitmodel}
